The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-28 article_final/main.tex
refined text in section 4

\paragraph{Schatten norms}
\citep{cesista2025schattenp} considered Schatten-$p$ norms:
\[\norm{\mM^t}_{S_p} = \left(\sum_{i=1}^{\min(m, n)}\sigma_i^p\right)^{1/p}\,,
\]
which produce the update
\[
\mX^{t+1} = \mX^t - \gamma_t \mU \frac{\diag\left(\sigma_1^{q-1},\dots, \sigma_{\min(m,n)}^{q-1}\right)}{\left(\sum_{i=1}^{\min(m,n)}{\sigma_i^q}\right)^{\frac{q-1}{q}}}\mV^\top
\]
where $p$ and $q$ satisfy $\frac{1}{p} + \frac{1}{q} = 1$. This formula recovers Neon when $p\rightarrow1$ (provided $\sigma_1 > \sigma_2$, which holds on real data), NSGD when $p=2$, and Muon when $p\rightarrow\infty$.

However, Schatten norms do not fill the rank gap: when $p > 1$, the update has full rank, while when $p=1$ (or $p \rightarrow 1$), it has rank one. Moreover, computing the update for $p \neq 1, 2, \infty$ appears to require knowing all $\sigma_i$, making the problem as hard as computing the full SVD.

@@

Deriving the update for arbitrary $k$: Since $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \citep{bhatia2013matrix}, p. 96), the goal is to reach $\max\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i, \sigma_1\}$. The Neon update $\Delta = \vu_1 \vv_1^\top$ achieves $\sigma_1$, while the scaled Muon update $\Delta = \frac{1}{k}\sum_{i=1}^{\min(m,n)}\vu_i \vv_i^\top$ achieves $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
\[
\mX^{t+1} = \mX^{t} - \gamma_t \vu_1 \vv_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\gamma_t}{k}\mU \mV^\top\,,
\]
depending on which better minimizes the linear function $L(\mX) \equiv F(\mX^t) + \mM^t(\mX - \mX^t)$.

The Ky Fan norms thus fail to close the rank gap: the resulting update is either rank-one or full-rank.

@@

While Schatten norms are closed under dualization and $\norm{\mM^t}_{S_p}^\dagger = \norm{\mM^t}_{S_q}$, for Ky Fan norms it is not generally the case. $k = 1$ and $k=\min(m, n)$ are exceptional: for $k=1$ the dual norm is $\normn{\cdot}$ with $k = \min(m, n)$ and for $k = \min(m, n)$ the dual norm is $\norms{\cdot}$ with $k=1$. For each other $k$, $\normkfk{\cdot} \neq \norm{\cdot}_{\text{KF-}k'}$ for any $k'$: $k'=\min(m, n)$ and $k'= 1$ correspond to the already discussed cases, while for other $k'$ one can change $\norm{\mM^t}_{\text{KF-}k'}$ by moving a small value between $\sigma_{k'}$ and $\sigma_{k'+1}$. During this change, $\norms{\mM^t}$ and $\normn{\mM^t}$ will remain constant, hence $\normkfk{\mM^t}$ will not change as well.

The key insight is to work with the duals of Ky Fan norms:

\begin{lemma}\label{lemma:ky_fan_update}
	When $\norm{\cdot} = \normkfk{\mM^t}^\dagger$, \equationref{eq:our_update} turns into:

	\begin{equation}\label{eq:ky_fan_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \sum_{i=1}^{k}\vu_i \vv_i^\top
	\end{equation}
\end{lemma}

\begin{proof}
Since $\norm{\cdot}^\dagger = \normkfk{\cdot}^{\dagger\dagger} = \normkfk{\cdot}$, the goal is to reach $\normkfk{\mM^t}$.

Let us note that $\Delta = \sum_{i=1}^{k} \vu_i \vv_i^\top$ delivers the value. Indeed,
\[
\<\mM^t, \Delta> =\<\mU \mSigma \mV^\top, \sum_{i=1}^{k} \vu_i \vv_i^\top> = \sum_{i,j=1}^{r, k}\<\vu_i \sigma_i \vv_i^\top, \vu_j \vv_j^\top> = \sum_{i=1}^{k}\sigma_i = \normkfk{\mM^t}\,,
\]
which completes the proof.
\end{proof}

These updates constitute the family of \emph{Fanions}, LMO-based algorithms under the $\normkfk{\mM^t}^\dagger$ norms. We call the algorithm for a particular $k$ \emph{Fanion-$k$}. This family elegantly bridges the rank gap with updates of any intermediate rank $k$.

\paragraph{Connection to existing algorithms}
The Fanion family unifies several known algorithms as special cases:
\begin{itemize}
	\item \textbf{Neon} is Fanion-1 (rank-one updates)
	\item \textbf{Muon} is Fanion-$\min\{m,n\}$ (full-rank updates)
	\item \textbf{Dion} (unsharded): The rank-$r$ Dion (Algorithm 1 from \citep{ahn2025dion}) without error feedback and without scaling of the update is actually Fanion-$r$ (see \citep{pethick2025understandingdion}, where Dion is written in a notation more similar to ours)
\end{itemize}

@@

Simply applying the norm dual to the Ky Fan $k$-norm, however, is not enough to provide a family of algorithms diverse enough for our purposes. Next, we consider linear combinations of algorithms from the LMO framework, which happen to be LMO-algorithms as well, as we show in the following paragraphs.

\subsection{General Case}
First, we need a well-known fact mentioned, for example, in \citep[Table 1]{yu2012arithmetic}. For the sake of completeness, we provide a proof here.

@@

@@

\begin{proof}
	Using \lemmaref{lemma:dual_to_conv_comb} and the fact $\norm{\cdot}^{\dagger \dagger} = \norm{\cdot}$, we can obtain general form of the unit ball in the $\norm{\cdot}$ norm: $\cB_{\norm{\cdot}} = \sum_{i = 1}^n \alpha_i \cB_{\norm{\cdot}_{(i)}}$. Thus, the linear function minimization objective on a $\norm{\cdot}$ norm ball can be transformed as follows:
	\[
	\argmin_{\mD \in \cB_{\norm{\cdot}}}\<\mM, \mD> = \argmin_{\mD_1 \in \alpha_i \cB_{\norm{\cdot}_{(1)}},\dots,\mD_n \in \alpha_n\cB_{\norm{\cdot}_{(n)}}}\<\mM, \sum_{i = 1}^n\mD_i> = \sum_{i = 1}^n \argmin_{D_i \in \cB_{\norm{\cdot}_{(i)}}}\<\mM, \mD_i>\,,
	\]
	where the last summation denotes the Minkowski sum. This immediately implies $\sum_{i = 1}^n \alpha_i \mathrm{LMO}_{i} \in \argmin_{\mD \in \cB_{\norm{\cdot}}}\<\mM, \mD>$, which proves the claim of the lemma.
\end{proof}

Applying it to optimization algorithms, we obtain the following corollary.

@@

\subsection{Frobeniusize Them: F-Muon and F-Neon}
Let us construct the concrete examples of algorithms given by linear combinations of LMO-algorithms. It follows from \corollaryref{corollary:lin_comb_alg} that those algorithms can also be viewed as LMO-algorithms.

Combining Fanions for various $k$ with another excellent LMO-algorithm, NSGD, we obtain the family of algorithms with updates defined by
\begin{equation}\label{eq:nsgd_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\frac{\mM^t}{\normf{\mM^t}}\right)\,.
\end{equation}
Recall that Fanion-$k$ update is governed by the norm dual to the Ky Fan-$k$ norm, and NSGD corresponds to the Frobenius norm, which is dual to itself. Thus, by \corollaryref{corollary:lin_comb_alg}, our new algorithm is an LMO-algorithm corresponding to the unit ball of the norm $\norm{\cdot}_{\mathrm{F-KF-k}}^\dagger$:
\begin{equation}\label{eq:fkfk_norm}
	\norm{\cdot}_{\mathrm{F-KF-k}} = \alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}\,.
\end{equation}
We name the derived family of algorithms \emph{F-Fanions}.

Two edge members of this family, with $k = 1$ and $k = \min\{m, n\}$ correspondingly, \emph{F-Neon} and \emph{F-Muon}, are of particular interest to us. A bit of information and visualizations related to the $\normfstar{\cdot} = \norm{\cdot}_{\mathrm{F-KF-1}}$ norm is presented in the appendix (see \equationref{eq:normfstardual}, \figureref{fig:fstardual}).

@@

Another algorithm we consider is the SignSGD. Mixing its update with that of Fanion-$k$, we obtain the update of \emph{S-Fanion-$k$}:
\begin{equation}\label{eq:sign_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\eta\sign(\mM^t)\right)\,,
\end{equation}
where $\eta$ is the special SignSGD learning rate coefficient.

The norm $\norm{\cdot}_{\mathrm{C-KF-k}}^\dagger$ that produces this update in the LMO framework is again given by \corollaryref{corollary:lin_comb_alg}:
\begin{equation}\label{eq:ckfk_norm}
	\norm{\cdot}_{\mathrm{C-KF-k}} = \alpha \normkfk{\cdot} + \frac{1-\alpha}{\eta}\normc{\cdot}^\dagger\,.
\end{equation}

@@

First, we compare F-Fanions on the following convex $L$-smooth problem:
\begin{equation}\label{eq:lls}
  F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\,
\end{equation}
where $\mX \in \Rmn$, $m=500$ and $n=500$ are typical dimensions of a neural network weight matrix, $\mS \in \Rmn$, $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed in the $(0, 1)$ interval. We set $\mS=0$, and elements of $\mX^0$ are instances of random value with distribution $\sim 0.1 \cN(0, 1)$.

@@

>>>>

\paragraph{Schatten norms}
\citep{cesista2025schattenp} considered Schatten-$p$ norms:
\[\norm{\mM^t}_{S_p} = \left(\sum_{i=1}^{\min(m, n)}\sigma_i^p\right)^{1/p}\,,
\]
which produce the update
\[
\mX^{t+1} = \mX^t - \gamma_t \mU \frac{\diag\left(\sigma_1^{q-1},\dots, \sigma_{\min(m,n)}^{q-1}\right)}{\left(\sum_{i=1}^{\min(m,n)}{\sigma_i^q}\right)^{\frac{q-1}{q}}}\mV^\top
\]
where $p$ and $q$ satisfy $p^{-1} + q^{-1} = 1$. This formula recovers Neon when $p\rightarrow1$ (provided $\sigma_1 > \sigma_2$, which holds on real data), NSGD when $p=2$, and Muon when $p\rightarrow\infty$.

However, Schatten norms do not fill the rank gap: when $p > 1$, the update has full rank, while when $p=1$, it has rank one. Moreover, computing the update for $p \neq 1, 2, \infty$ appears to require knowing all $\sigma_i$, making the problem as hard as computing the full SVD.

@@

To derive the update formula for arbitrary $k$, we use the expression for the norm dual to the Ky Fan $k$-norm (see, for instance \citep{bhatia2013matrix}, p. 96):
\[
    \normkfk{\cdot}^\dagger = \max\left\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\right\}\,.
\]
According to \equationref{eq:our_update}, we need to find $\mD$ such that $\normkfk{\mD} = 1$ and
\[
\<\mM^t, \mD> = \max\left\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i,\ \sigma_1\right\}\,.
\]
The Neon update $\mD = \vu_1 \vv_1^\top$ achieves $\sigma_1$, while the scaled Muon update $\mD = \frac{1}{k}\sum_{i=1}^{\min(m,n)}\vu_i \vv_i^\top$ achieves $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
\[
\mX^{t+1} = \mX^{t} - \gamma_t \vu_1 \vv_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\gamma_t}{k}\mU \mV^\top\,,
\]
depending on which of those two expressions achieves smaller residual on the linear function $L(\mX) := F(\mX^t) + \<\mM^t, \mX - \mX^t>$. The Ky Fan norms thus fail to close the rank gap: the resulting update is either rank-one or full-rank.

@@

Unlike Schatten norms, which satisfy $\norm{\mM^t}_{S_p}^\dagger = \norm{\mM^t}_{S_q}$ (for $p^{-1} + q^{-1} = 1$) and are thus closed under dualization, Ky Fan norms are generally not. Only two exceptional cases exist: the dual of the Ky Fan $1$-norm $\norm{\cdot}_{\text{KF-}1}$ (the spectral norm) is the Ky Fan $\min(m,n)$-norm $\norm{\cdot}_{\text{KF-}\min(m,n)}$ (the nuclear norm), and vice versa. Thus working with duals of Ky Fan norms can open up new possibilities:

\begin{lemma}\label{lemma:ky_fan_update}
	When $\norm{\cdot} = \normkfk{\mM^t}^\dagger$, \equationref{eq:our_update} turns into:
	\begin{equation}\label{eq:ky_fan_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \sum_{i=1}^{k}\vu_i \vv_i^\top\,.
	\end{equation}
\end{lemma}

\begin{proof}
Since $\norm{\cdot}^\dagger = \normkfk{\cdot}^{\dagger\dagger} = \normkfk{\cdot}$, the goal is to reach $\<\mM^t, \mD> = \normkfk{\mM^t}$ in \equationref{eq:our_update}. Note that $\mD = \sum_{i=1}^{k} \vu_i \vv_i^\top$ attains this value. Indeed,
\[
\<\mM^t, \mD> =\<\mU \mSigma \mV^\top, \sum_{i=1}^{k} \vu_i \vv_i^\top> = \sum_{i,j=1}^{r, k}\<\vu_i \sigma_i \vv_i^\top, \vu_j \vv_j^\top> = \sum_{i=1}^{k}\sigma_i = \normkfk{\mM^t}\,,
\]
which completes the proof.
\end{proof}

These updates define the \emph{Fanion} family of LMO-based algorithms, each operating under a norm $\normkfk{\mM^t}^\dagger$. We denote the algorithm for a particular $k$ as \emph{Fanion-$k$}. This family elegantly bridges the rank gap, providing updates of any intermediate rank $k$.

\paragraph{Connection to existing algorithms}
The Fanion family unifies several known algorithms:
\begin{itemize}
	\item \textbf{Neon} is Fanion-1 (rank-one updates)
	\item \textbf{Muon} is Fanion-$\min\{m,n\}$ (full-rank updates)
	\item \textbf{Dion} (unsharded): The rank-$r$ Dion (Algorithm 1 from \citep{ahn2025dion}) without error feedback and without scaling of the update is actually Fanion-$r$ (see \citep{pethick2025understandingdion}, where Dion is written in a notation more similar to ours)
\end{itemize}

@@

Simply applying norms dual to Ky Fan $k$-norms, however, does not provide sufficient algorithmic diversity for our purposes. We now consider linear combinations of LMO-based algorithms and show that these combinations are themselves LMO-algorithms.

\subsection{General Case}
We begin with a well-known result on dual norms (see, e.g., \citep[Table 1]{yu2012arithmetic}).

@@

A proof is provided in the appendix (see \sectionref{sec:proof_dual_conv_comb}).

@@

\begin{proof}
	Using \lemmaref{lemma:dual_to_conv_comb} and the biduality property $\norm{\cdot}^{\dagger \dagger} = \norm{\cdot}$, we obtain the unit ball representation: $\cB_{\norm{\cdot}} = \sum_{i = 1}^n \alpha_i \cB_{\norm{\cdot}_{(i)}}$. The linear minimization problem over this ball can thus be transformed as follows:
	\[
	\argmin_{\mD \in \cB_{\norm{\cdot}}}\<\mM, \mD> = \argmin_{\mD_1 \in \alpha_i \cB_{\norm{\cdot}_{(1)}},\dots,\mD_n \in \alpha_n\cB_{\norm{\cdot}_{(n)}}}\<\mM, \sum_{i = 1}^n\mD_i> = \sum_{i = 1}^n \argmin_{D_i \in \cB_{\norm{\cdot}_{(i)}}}\<\mM, \mD_i>\,,
	\]
	where the last summation denotes the Minkowski sum. This immediately implies $\sum_{i = 1}^n \alpha_i \mathrm{LMO}_{i} \in \argmin_{\mD \in \cB_{\norm{\cdot}}}\<\mM, \mD>$, completing the proof.
\end{proof}

Applying this result to optimization algorithms yields the following corollary.

@@

\subsection{Frobeniusize Them: F-Muon and F-Neon}
We now construct concrete examples of algorithms obtained via linear combinations of LMO-algorithms. By \corollaryref{corollary:lin_comb_alg}, these combinations are themselves LMO-algorithms.

Combining Fanion-$k$ with NSGD yields a family of algorithms with updates
\begin{equation}\label{eq:nsgd_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\frac{\mM^t}{\normf{\mM^t}}\right)\,.
\end{equation}
Recall that Fanion-$k$ operates under the dual to the Ky Fan $k$-norm, while NSGD operates under the self-dual Frobenius norm. By \corollaryref{corollary:lin_comb_alg}, this combination defines an LMO-algorithm with norm $\norm{\cdot}_{\mathrm{F-KF-k}}^\dagger$, where
\begin{equation}\label{eq:fkfk_norm}
	\norm{\cdot}_{\mathrm{F-KF-k}} = \alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}\,.
\end{equation}
We call this family \emph{F-Fanions}.

The extremal members of this family are \emph{F-Neon} (with $k = 1$) and \emph{F-Muon} (with $k = \min\{m, n\}$). Additional information and visualizations for the $\normfstar{\cdot} = \norm{\cdot}_{\mathrm{F-KF-1}}$ norm appear in the appendix (see \equationref{eq:normfstardual}, \figureref{fig:fstardual}).

@@

Similarly, combining Fanion-$k$ with SignSGD yields \emph{S-Fanion-$k$} with update
\begin{equation}\label{eq:sign_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\eta\sign(\mM^t)\right)\,,
\end{equation}
where $\eta$ is a scaling coefficient specific to SignSGD.

By \corollaryref{corollary:lin_comb_alg}, this defines an LMO-algorithm with norm $\norm{\cdot}_{\mathrm{C-KF-k}}^\dagger$, where
\begin{equation}\label{eq:ckfk_norm}
	\norm{\cdot}_{\mathrm{C-KF-k}} = \alpha \normkfk{\cdot} + \frac{1-\alpha}{\eta}\normc{\cdot}^\dagger\,.
\end{equation}

The extremal members of this family are \emph{S-Neon} (with $k = 1$) and \emph{S-Muon} (with $k = \min\{m, n\}$).

@@

First, we compare F-Fanions on the following convex $L$-smooth problem:
\begin{equation}\label{eq:lls}
  F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\,
\end{equation}
where $\mX \in \Rmn$, $m=500$ and $n=500$ are typical dimensions of a neural network weight matrix, $\mS \in \Rmn$, $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed in the $(0, 1)$ interval. We set $\mS=0$, and $\mX^0 \sim 0.1 \cN(0, \mI)$.

@@

\section{Proof of Lemma on Dual of Convex Combinations}\label{sec:proof_dual_conv_comb}

We provide here the proof of \lemmaref{lemma:dual_to_conv_comb}.
